Beliefs about the development of mental life

### 2025-02-27 paper/cover.tex
de-anonymize

>>>>

    \indent Weisman, King, and Humphreys were affiliated with the Department of Psychology at Stanford University at the time of data collection. Weisman was affiliated with the Department of Psychology at the University of California, Riverside, at the time of submission, and is now affiliated with the Department of Psychology at Harvard University. King and Humphreys are now affiliated with the Department of Psychology and Human Development at Vanderbilt University.

### 2025-02-27 paper/paper.Rmd
de-anonymize

---
# title: "Beliefs about the development of mental life"
# author: 
#   - "Kara Weisman\\newline University of California, Riverside\\newline ORCID: https://orcid.org/0000-0002-3706-7554"
#   - "Lucy S. King\\newline Vanderbilt University\\newline ORCID: http://orcid.org/0000-0002-2552-0614"
#   - "Kathryn Humphreys\\newline Vanderbilt University\\newline ORCID: https://orcid.org/0000-0002-5715-6597"
# output:
#   pdf_document:  default
#   # word_document:
#     # reference_docx: word-styles-ref.docx
output:
  bookdown::pdf_document2:
    includes:
      in_header: header.tex
      before_body: cover.tex
    toc: false
    number_sections: false
bibliography: ../supplement/references.bib
csl: ../supplement/apa.csl
figsintext: yes
header-includes:
- \usepackage{lineno}
- \linenumbers
- \modulolinenumbers[5]
- \usepackage{float}
---

@@

Methods, inclusion/exclusion criteria, and analyses marked as "preregistered" were preregistered on the Open Science Framework (OSF) website (Study 1: https://osf.io/e6ajh; Study 2: https://osf.io/j72dg; Study 3: https://osf.io/xh8ce). All studies were approved by the Research Compliance Office, Human Subjects Panel (IRB) at Stanford University. For extended descriptions of methods and results, see Supplemental Materials.

@@

Participants completed three trials in which they assessed the mental lives of children at three target ages: birth, 9 months, and 5 years. On each trial, they were shown two representative photographs of children at the target age (labeled "newborns," "9-month-olds," or "5-year-olds"), and asked to answer the following question for 60 capacities: "To what extent is a [newborn/9-month-old/5-year-old] capable of [this capacity]?" See Supplemental Materials for details about the selection of these photographs. These photographs (and all materials used in these studies) are available in the following OSF repository: https://osf.io/xrznd/ (DOI: 10.17605/OSF.IO/XRZND). Participants responded on a sliding scale from 0 (labeled "not at all capable") to 100 ("completely capable").

@@

@@

@@

@@

First, are the beliefs and intuitive theories documented here accurate with respect to the current scientific understanding of human development? Comparing participants' intuitions to empirical findings from scientific studies of child development has illuminated the ways in which lay people's reasoning about cognitive development [@berent.etal2019; @wang.feigenson2019] and about emotions [@berent.etal2020] has proved a useful way of illuminating biases and misconceptions in laypeople's beliefs about innateness. Finding creative ways to determine the ground truth of the development of the range of capacities included in the current studies—many of which are phenomenal experiences—will be a particular challenge to this endeavor, but would be both deeply fascinating and of practical interest to predicting downstream consequences for children. All else being equal, we would predict that caregivers with more accurate representations of children's developing minds would make more accurate inferences about their own children and interact with their children in ways that are more developmentally appropriate and better-suited to scaffolding the child's development [see, e.g., @list.etal2021].

>>>>

---
output:
  bookdown::pdf_document2:
    includes:
      in_header: header.tex
      before_body: cover.tex
    toc: false
    number_sections: false
bibliography: ../supplement/references.bib
csl: ../supplement/apa.csl
figsintext: yes
header-includes:
- \usepackage{lineno}
- \linenumbers
- \modulolinenumbers[5]
- \usepackage{float}
- \usepackage{url}
---

@@

Methods, inclusion/exclusion criteria, and analyses marked as "preregistered" were preregistered on the Open Science Framework (OSF) website (Study 1: \url{https://osf.io/e6ajh}; Study 2: \url{https://osf.io/j72dg}; Study 3: \url{https://osf.io/xh8ce)}). All studies were approved by the Stanford University Research Compliance Office (Human Subjects IRB). For extended descriptions of methods and results, see Supplemental Materials. Note: scripts for all analyses, figures, and tables are available in the .Rmd file that generated the fully reproducible Supplemental Materials document (supplement-main.Rmd), which can be found in the following GitHub repository: \url{https://github.com/kgweisman/baby_mental_life_ms}.

@@

Participants completed three trials in which they assessed the mental lives of children at three target ages: birth, 9 months, and 5 years. On each trial, they were shown two representative photographs of children at the target age (labeled "newborns," "9-month-olds," or "5-year-olds"), and asked to answer the following question for 60 capacities: "To what extent is a [newborn/9-month-old/5-year-old] capable of [this capacity]?" See Supplemental Materials for details about the selection of these photographs. These photographs (and all materials used in these studies) are available in the following OSF repository: \url{https://osf.io/xrznd}. Participants responded on a sliding scale from 0 (labeled "not at all capable") to 100 ("completely capable").

@@

```{r fig04, fig.cap="Network graph summarizing the importance of each developmental mechanism (along the bottom) for each capacity (along the top) among US adults (Study 3). Capacities are sorted and color-coded by domain (see Figures 2-3); developmental mechanisms are arranged in the fixed order of presentation (roughly, from what we perceived to be the most innate or biological mechanisms of development, to what we perceived to be the most learned or social). See main text for the full text of each mechanism.", fig.width = 8, fig.asp = 0.8}
knitr::include_graphics("../outputs/fig04.pdf")
```

@@

Related to the issue of the cultural specificity of our findings is the issue of the specificity of our stimuli. In selecting photos for each target age, we aimed to have photos of the same (or very similar-looking) children at all target ages so as to emphasize the question of the development of mental capacities over the course of an individual child’s early life. We chose to illustrate each target age with two photos in order to emphasize that we were interested in children’s mental capacities in general, rather than the experiences of a particular child in a particular moment. We included one boy and one girl at each target age in order to generalize across potential perceived sex/gender differences in children’s early mental life. We used photographs of two (apparently) White, non-Hispanic children, for three reasons: (1) in order to avoid introducing spurious differences across target ages by confounding a target’s age and race/ethnicity; (2) in order to match the anticipated race/ethnicity of the majority of participants on MTurk; and (3) because of the limited set of high-resolution photographs of individual children of color at different points of development available to us. Conducting similar studies of adults’ perceptions of the development of mental life with more ethnically and racially diverse participants, and examining the effects of sex, gender, race, and ethnicity on adults’ perceptions of children’s mental life, will be important next steps in this line of research.

On a similar note, although our instructions emphasized to participants that we were interested in their assessment of children in general, rather than the depicted children in particular, it is nearly inevitable that the particular photographs used as stimuli—not just race and ethnicity, but also facial expressions, attractiveness, cues to social class, and so on—would guide participants’ responses. Future research might employ a variety of approaches to address this concern, such as constructing a much larger and more diverse pool of visual stimuli and varying the subset of stimuli provided across participants; blurring children’s facial expressions; blurring or removing the background of the photographs; or removing all visual stimuli from the study (likely more appropriate for parents of young children than for a general population study).

@@

Finally, while our current analytical approach identifies key patterns in these developmental trajectories, future work could benefit from employing functional data analysis methods, such as functional principal components analysis (fPCA), which might capture smooth, continuous variations in temporal patterns and better account for the inherent time-dependency in developmental data [@ramsay.silverman2002]. Such an approach might identify clusters of mental capacities that hang together not only in terms of overall covariance, but also in terms of the perceived timing and rate of development, or in terms of other temporal dynamics that our current approach might not detect. Employing such techniques would be an exciting next step in this line of research.

@@

First, are the beliefs and intuitive theories documented here accurate with respect to the current scientific understanding of human development? Comparing participants' intuitions to empirical findings from scientific studies of child development has illuminated the ways in which lay people's reasoning about cognitive development [@berent.etal2019; @wang.feigenson2019] and about emotions [@berent.etal2020] has proved a useful way of illuminating biases and misconceptions in laypeople's beliefs about innateness. The primary goal of the current studies was to characterize how US adults think about development, rather than to generate precise estimates of the age at which lay people consider each of these capacities to “come online”; nonetheless, each of these studies provides estimates of participants’ beliefs about children’s capacities at different ages, and the collection of these studies provides some indication of how stable these estimates are across different methods of administration; see Supplemental Materials for direct comparisons of developmental trajectories across studies at the level of the individual capacity (Figure S9) and the domain (Figure S10). Finding creative ways to determine the ground truth of the development of the range of capacities included in the current studies—many of which are phenomenal experiences—will be a particular challenge to this endeavor, but would be both deeply fascinating and of practical interest to predicting downstream consequences for children. All else being equal, we would predict that caregivers with more accurate representations of children's developing minds would make more accurate inferences about their own children and interact with their children in ways that are more developmentally appropriate and better-suited to scaffolding the child's development [see, e.g., @list.etal2021].

### 2025-02-27 supplement/supplement-main.Rmd
de-anonymize

---
# title: "Beliefs about the development of mental life"
# subtitle: "Supplemental Materials"
# author: "Kara Weisman, Lucy S. King, and Kathryn Humphreys"
# date: "2024-12-30"
output:
  bookdown::pdf_document2:
    includes:
      in_header: header.tex
      before_body: cover.tex
    latex_engine: xelatex
    fig_caption: yes
    toc: true
    toc_depth: 2
    number_sections: false
  #   reference_docx: word-styles-ref.docx
  # word_document:
bibliography: "references.bib"
csl: apa.csl
# cls: article.cls
figsintext: yes
header-includes: 
   - \usepackage{lineno}
   - \linenumbers
   - \modulolinenumbers[5]
   - \usepackage{float}
   - \usepackage{pdflscape}
tables: yes
---

@@

Note: This is a fully reproducible manuscript, written in RMarkdown using the "knitr" package for R [@xie2021]. The full, editable script for all analyses, figures, and tables (here and in the main text) is available in the .Rmd file that generated this document, which can be found in the following GitHub repository: https://github.com/kgweisman/baby_mental_life_ms.

@@

Methods, inclusion/exclusion criteria, and analyses marked as "preregistered" were preregistered on the Open Science Framework (OSF) website (Study 1: https://osf.io/e6ajh; Study 2: https://osf.io/j72dg; Study 3: https://osf.io/xh8ce).

All materials used in these studies are available in the following OSF repository: https://osf.io/xrznd/ (DOI: 10.17605/OSF.IO/XRZND).

All studies were approved by the Institution Review Board for Human Subjects at Stanford University.

@@

```{r figS06 code, fig.width = 10, fig.asp = 1.2}
figS06 <- d123_gam %>%
  filter(target_ord %in% c("newborns", "9-month-olds", "5-year-olds")) %>%
  filter(capacity %in%
           c(full_join(d2_gam %>% distinct(capacity),
                       d3_gam %>% distinct(capacity)) %>%
               distinct(capacity) %>%
               pull(capacity) %>%
               as.character())) %>%
  complete(
    capacity, nesting(study, target, target_year, target_num, target_ord),
    fill = list(response = -1)) %>% # introduce -1 to align studies across panels
  select(-factor) %>%
  left_join(d2_gam %>% distinct(capacity, factor)) %>%
  left_join(d2_scales %>% select(-loading)) %>%
  mutate(capacity_text = str_wrap(gsub("_", " ", capacity), 40)) %>%
  ggplot(aes(x = target_ord, y = response, 
             color = factor)) +
  facet_wrap(facets = vars(reorder(capacity_text, as.numeric(order))), nrow = 4) +
  geom_point(
    aes(group = study),
    size = 0.02,
    alpha = 0.6,
    position = position_jitterdodge(
      dodge.width = 0.8,
      jitter.width = 0.6,
      jitter.height = 0)) +
  scale_color_viridis_d(end = 0.9, option = "viridis") +
  geom_path(
    data = . %>%
      group_by(study, target, target_year, target_num, target_ord, 
               factor, capacity, capacity_text, order) %>%
      summarise(mean = mean(response, na.rm = T)) %>%
      ungroup(),
    aes(y = mean, group = study),
    color = "black", lty = 2, 
    position = position_dodge(width = 0.8)) +
  geom_pointrange(
    data = . %>%
      group_by(study, target, target_year, target_num, target_ord, 
               factor, capacity, capacity_text, order) %>%
      multi_boot_standard(col = "response") %>%
      ungroup(),
    aes(y = mean, ymin = ci_lower, ymax = ci_upper, 
        group = study, fill = study),
    color = "black", shape = 21, fatten = 6,
    position = position_dodge(width = 0.8)) +
  geom_text(
    data = . %>%
      group_by(study, target, target_year, target_num, target_ord, 
               factor, capacity, capacity_text, order) %>%
      summarise(mean = mean(response, na.rm = T)) %>%
      ungroup(),
    aes(y = mean, group = study, label = gsub("Study ", "", study)),
    color = "white", size = 2.5, fontface = "bold", hjust = 0.5, vjust = 0.5,
    position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("#9801a3", "#a3012f", "#a36501")) +
  scale_y_continuous(
    limits = c(-0.05, NA), # exclude the specious "-1" responses introduced above
    breaks = seq(0, 100, 25),
    labels = seq(0, 100, 25)) +
  scale_x_discrete(labels = c("newborns", "9 months", "5 years")) +
  theme_minimal() +
  theme(panel.background = element_rect(fill = NA), 
        strip.text = element_text(vjust = 0),
        legend.position = "top") +
        # axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
        # plot.margin = unit(c(0.5, 0.5, 0.5, 1.1), "cm")) +
  labs(x = "Target age",
       y = "Mean response\n0 = not at all capable, 100 = completely capable",
       color = "Domain:",
       fill = "Study:") +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2)))
figS06

if (resave_plots == T | !file.exists("../outputs/figS06.pdf")) {
  ggsave("../outputs/figS06.svg")
  ggsave("../outputs/figS06.pdf")
}
```

```{r figS06, fig.cap="Comparison of capacity ratings across studies (Studies 1-3). This visualization includes responses for each of the capacities and target ages held in common across at least two of our three studies. Lighter colorful represent individual participants’ responses or scores, larger points correspond to mean responses or scores across the sample, and error bars are bootstrapped 95% confidence intervals.", include = T}
knitr::include_graphics("../outputs/figS06.pdf")
```

>>>>

---
output:
  bookdown::pdf_document2:
    includes:
      in_header: header.tex
      before_body: cover.tex
    latex_engine: xelatex
    fig_caption: yes
    toc: true
    toc_depth: 2
    number_sections: false
  #   reference_docx: word-styles-ref.docx
  # word_document:
bibliography: "references.bib"
csl: apa.csl
# cls: article.cls
figsintext: yes
header-includes: 
   - \usepackage{lineno}
   - \linenumbers
   - \modulolinenumbers[5]
   - \usepackage{float}
   - \usepackage{url}
   - \usepackage{pdflscape}
   - \makeatletter
   - \renewcommand{\fnum@figure}{Figure S\thefigure}
   - \makeatother
tables: yes
---

@@

Note: This is a fully reproducible manuscript, written in RMarkdown using the "knitr" and "bookdown" packages for R [@xie2016; @xie2021; @xie2024]. The full, editable script for all analyses, figures, and tables (here and in the main text) is available in the .Rmd file that generated this document (supplement-main.Rmd), which can be found in the following GitHub repository: \url{https://github.com/kgweisman/baby_mental_life_ms}.

@@

Methods, inclusion/exclusion criteria, and analyses marked as "preregistered" were preregistered on the Open Science Framework (OSF) website (Study 1: \url{https://osf.io/e6ajh}; Study 2: \url{https://osf.io/j72dg}; Study 3: \url{https://osf.io/xh8ce)}).

All materials used in these studies are available in the following OSF repository: \url{https://osf.io/xrznd}.

All studies were approved by the Stanford University Research Compliance Office (Human Subjects IRB).

@@

```{r figS13 code, fig.width = 10, fig.asp = 1.2}
figS13 <- d123_gam %>%
  filter(target_ord %in% c("newborns", "9-month-olds", "5-year-olds")) %>%
  filter(capacity %in%
           c(full_join(d2_gam %>% distinct(capacity),
                       d3_gam %>% distinct(capacity)) %>%
               distinct(capacity) %>%
               pull(capacity) %>%
               as.character())) %>%
  complete(
    capacity, nesting(study, target, target_year, target_num, target_ord),
    fill = list(response = -1)) %>% # introduce -1 to align studies across panels
  select(-factor) %>%
  left_join(d2_gam %>% distinct(capacity, factor)) %>%
  left_join(d2_scales %>% select(-loading)) %>%
  mutate(capacity_text = str_wrap(gsub("_", " ", capacity), 40)) %>%
  ggplot(aes(x = target_ord, y = response, 
             color = factor)) +
  facet_wrap(facets = vars(reorder(capacity_text, as.numeric(order))), nrow = 4) +
  geom_point(
    aes(group = study),
    size = 0.02,
    alpha = 0.6,
    position = position_jitterdodge(
      dodge.width = 0.8,
      jitter.width = 0.6,
      jitter.height = 0)) +
  scale_color_viridis_d(end = 0.9, option = "viridis") +
  geom_path(
    data = . %>%
      group_by(study, target, target_year, target_num, target_ord, 
               factor, capacity, capacity_text, order) %>%
      summarise(mean = mean(response, na.rm = T)) %>%
      ungroup(),
    aes(y = mean, group = study),
    color = "black", lty = 2, 
    position = position_dodge(width = 0.8)) +
  geom_pointrange(
    data = . %>%
      group_by(study, target, target_year, target_num, target_ord, 
               factor, capacity, capacity_text, order) %>%
      multi_boot_standard(col = "response") %>%
      ungroup(),
    aes(y = mean, ymin = ci_lower, ymax = ci_upper, 
        group = study, fill = study),
    color = "black", shape = 21, fatten = 6,
    position = position_dodge(width = 0.8)) +
  geom_text(
    data = . %>%
      group_by(study, target, target_year, target_num, target_ord, 
               factor, capacity, capacity_text, order) %>%
      summarise(mean = mean(response, na.rm = T)) %>%
      ungroup(),
    aes(y = mean, group = study, label = gsub("Study ", "", study)),
    color = "white", size = 2.5, fontface = "bold", hjust = 0.5, vjust = 0.5,
    position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("#9801a3", "#a3012f", "#a36501")) +
  scale_y_continuous(
    limits = c(-0.05, NA), # exclude the specious "-1" responses introduced above
    breaks = seq(0, 100, 25),
    labels = seq(0, 100, 25)) +
  scale_x_discrete(labels = c("newborns", "9 months", "5 years")) +
  theme_minimal() +
  theme(panel.background = element_rect(fill = NA), 
        strip.text = element_text(vjust = 0),
        legend.position = "top") +
        # axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
        # plot.margin = unit(c(0.5, 0.5, 0.5, 1.1), "cm")) +
  labs(x = "Target age",
       y = "Mean response\n0 = not at all capable, 100 = completely capable",
       color = "Domain:",
       fill = "Study:") +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2)))
figS13

if (resave_plots == T | !file.exists("../outputs/figS13.pdf")) {
  ggsave("../outputs/figS13.jpg")
  ggsave("../outputs/figS13.svg")
  ggsave("../outputs/figS13.pdf")
}
```

```{r figS13, fig.cap="Comparison of capacity ratings across studies (Studies 1-3). This visualization includes responses for each of the capacities and target ages held in common across at least two of our three studies. Lighter colorful points represent individual participants’ responses, larger points correspond to mean responses across the sample, and error bars are bootstrapped 95% confidence intervals.", include = T}
knitr::include_graphics("../outputs/figS13.pdf")
```

```{r}
d23_gam_pred <- d2_gam_r1_pred %>%
  mutate(study = "Study 2") %>%
  full_join(d3_gam_r1_pred %>% mutate(study = "Study 3")) %>%
  mutate(study = factor(study, levels = c("Study 1", "Study 2", "Study 3")))
```

```{r figS14 code, fig.width = 10, fig.asp = 0.5}
figS14 <- d23_gam_pred %>%
  full_join(
    d1_gam %>%
      mutate(study = "Study 1") %>%
      group_by(study, ResponseId, factor, target, target_year) %>%
      summarise(score = mean(response/100, na.rm = T))) %>%
  ggplot(aes(x = target_year, y = score, 
             color = study, fill = study)) +
  facet_grid(cols = vars(factor)) +
  geom_jitter(height = 0, width = 0.1, size = 0.01, alpha = 0.05) +
  geom_smooth(aes(y = pred_score, color = study, fill = study), 
              fill = NA, size = 0.8, lty = 1, 
              show.legend = F) +
  geom_path(
    data = . %>%
      filter(study == "Study 1") %>%
      # filter(target_year %in% c(0, 9/12, 5)) %>%
      group_by(study, target, target_year,
               factor) %>%
      summarise(mean = mean(score, na.rm = T)) %>%
      ungroup(),
    aes(y = mean, group = study),
    color = "#1b9e77", lty = 2) +
  geom_pointrange(
    data = . %>%
      # filter(study == "Study 1") %>%
      filter(target_year %in% c(0, 9/12, 5)) %>%
      group_by(study, target, target_year,
               factor) %>%
      multi_boot_standard(col = "score", na.rm = T) %>%
      ungroup(),
    aes(y = mean, ymin = ci_lower, ymax = ci_upper,
        group = study, fill = study),
    # color = "#1b9e77", 
    shape = 21, fatten = 2,
    show.legend = F) +
  # geom_text(
  #   data = . %>%
  #     group_by(study, target, target_year, target_num, target_ord, 
  #              factor) %>%
  #     summarise(mean = mean(score, na.rm = T)) %>%
  #     ungroup(),
  #   aes(y = mean, group = study, label = gsub("Study ", "", study)),
  #   color = "white", size = 2.5, fontface = "bold", hjust = 0.5, vjust = 0.5,
  #   position = position_dodge(width = 0.8)) +
  # scale_fill_manual(values = c("#9801a3", "#a3012f", "#a36501"), 
  #                   aesthetics = c("color", "fill")) +
  scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) +
  scale_y_continuous(
    breaks = seq(0, 1, 0.25),
    labels = seq(0, 100, 25)) +
  # scale_x_discrete(labels = c("newborns", "9 months", "5 years")) +
  # scale_x_discrete(labels = levels(d123_gam$target_ord)) +
  # scale_x_continuous(breaks = seq(0, 5, 1)) +
  theme_minimal() +
  theme(panel.background = element_rect(fill = NA), 
        strip.text = element_text(vjust = 0),
        legend.position = "top") +
        # axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
        # plot.margin = unit(c(0.5, 0.5, 0.5, 1.1), "cm")) +
  labs(x = "Target age (years)",
       y = "Mean score across all capacities in domain\n0 = not at all capable, 100 = completely capable",
       # caption = "Error bars are bootstrapped 95% CIs; solid lines are estimates from GAMs (Studies 1 and 2)",
       color = "Study:",
       fill = "Study:") +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2)))

figS14

if (resave_plots == T | !file.exists("../outputs/figS14.pdf")) {
  ggsave("../outputs/figS14.jpg")
  ggsave("../outputs/figS14.svg")
  ggsave("../outputs/figS14.pdf")
}
```

```{r figS14, fig.cap="Comparison of mean scores across studies (Studies 1-3). These scores incorporate responses to all of the capacities and target ages included in each study; this varies across studies. Lighter points represent individual participants’ scores, larger points correspond to mean scores across the sample, and error bars are bootstrapped 95% confidence intervals. The dotted line for Study 1 connects mean scores across target ages; the solid lines for Studies 2-3 reflect predictions from study-specific GAMs.", include = T}
knitr::include_graphics("../outputs/figS14.pdf")
```
